---
title: "Multivariate Normal Distribution"
---

## Bivariate Normal

### Correlation 0

```{r, echo = FALSE, message = FALSE}
library(rgl)
library(mvtnorm)
library(plyr)
library(dplyr)
library(ggplot2)
```

```{r, echo = FALSE, message = FALSE}
X1 <- unexpanded_X <- seq(from = -3, to = 3, length = 101)
X2 <- unexpanded_Y <- seq(from = -3, to = 3, length = 101)
plot_df <- as.data.frame(expand.grid(X = unexpanded_X, Y = unexpanded_Y))
plot_df_0 <- plot_df_0.5 <- plot_df_0.9 <- plot_df

plot_df_0$joint_density <- dmvnorm(plot_df, sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2))
plot_df_0.5$joint_density <- dmvnorm(plot_df, sigma = matrix(c(1, -0.5, -0.5, 1), nrow = 2, ncol = 2))
plot_df_0.9$joint_density <- dmvnorm(plot_df, sigma = matrix(c(1, 0.9, 0.9, 1), nrow = 2, ncol = 2))

color_n <- 1000 # number of colors used

joint_density_lim <- range(c(plot_df_0$joint_density, plot_df_0.5$joint_density, plot_df_0.9$joint_density))
joint_density_range <- joint_density_lim[2] - joint_density_lim[1]
joint_density_colorlut <- rev(rainbow(1000, start = 0, end = 4/6)) # height color lookup table
joint_density_col_0 <- joint_density_colorlut[ floor(color_n * (plot_df_0$joint_density - joint_density_lim[1])/joint_density_range) + 1 ]
joint_density_col_0.5 <- joint_density_colorlut[ floor(color_n * (plot_df_0.5$joint_density - joint_density_lim[1])/joint_density_range) + 1 ]
joint_density_col_0.9 <- joint_density_colorlut[ floor(color_n * (plot_df_0.9$joint_density - joint_density_lim[1])/joint_density_range) + 1 ]

junk <- open3d()
z_max <- joint_density_lim[2]
plot3d(X1, X2, xlim=c(-3, 3), ylim=c(-3, 3), zlim=c(0, z_max), zlab="f(x1, x2)", xlab = "x1", ylab = "x2", mouseMode = "zAxis", type = "s")
surface3d(X1, X2, plot_df_0$joint_density, alpha = 0.3, col = joint_density_col_0)

#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "plot_bivarnorm0")
```

```{r, echo = FALSE}
ggplot(plot_df_0, aes(x=X, y=Y, z=joint_density)) + 
    geom_contour() +
    coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), ratio = 1) +
    xlab("x1") +
    ylab("x2")
```

### Correlation -0.5

```{r, echo = FALSE, message = FALSE}
junk <- open3d()
z_max <- joint_density_lim[2]
plot3d(X1, X2, xlim=c(-3, 3), ylim=c(-3, 3), zlim=c(0, z_max), zlab="f(x1, x2)", xlab = "x1", ylab = "x2", mouseMode = "zAxis", type = "s")
surface3d(X1, X2, plot_df_0.5$joint_density, alpha = 0.3, col = joint_density_col_0.5)

#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "plot_bivarnorm0.5")
```

```{r, echo = FALSE}
ggplot(plot_df_0.5, aes(x=X, y=Y, z=joint_density)) + 
    geom_contour() +
    coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), ratio = 1) +
    xlab("x1") +
    ylab("x2")
```

### Correlation 0.9

```{r, echo = FALSE, message = FALSE}
junk <- open3d()
z_max <- joint_density_lim[2]
plot3d(X1, X2, xlim=c(-3, 3), ylim=c(-3, 3), zlim=c(0, z_max), zlab="f(x1, x2)", xlab = "x1", ylab = "x2", mouseMode = "zAxis", type = "s")
surface3d(X1, X2, plot_df_0.9$joint_density, alpha = 0.3, col = joint_density_col_0.9)

#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "plot_bivarnorm0.9")
```

```{r, echo = FALSE}
ggplot(plot_df_0.9, aes(x=X, y=Y, z=joint_density)) + 
    geom_contour() +
    coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), ratio = 1) +
    xlab("x1") +
    ylab("x2")
```

### Probabilities by Integration

$$P(0.5 \leq X_1 \leq 1.5 \text{ and } 0 \leq X_2 \leq 1) = \int_{0.5}^{1.5} \int_{0}^1 f(x_1, x_2 | \mathbf{\mu}, \Sigma) \, d x_2 d x_1$$

```{r, echo = FALSE}
x1_lower <- 0.5
x1_upper <- 1.5
x2_lower <- 0
x2_upper <- 1
X1_solid <- unexpanded_X_solid <- seq(from = x1_lower, to = x1_upper, length = 101)
X2_solid <- unexpanded_Y_solid <- seq(from = x2_lower, to = x2_upper, length = 101)
plot_df_solid <- as.data.frame(expand.grid(X = unexpanded_X_solid, Y = unexpanded_Y_solid))
plot_df_0_solid <- plot_df_solid

plot_df_0_solid$joint_density <- dmvnorm(plot_df_solid, sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2))

joint_density_lim <- range(plot_df_0$joint_density)
joint_density_range <- joint_density_lim[2] - joint_density_lim[1]
joint_density_colorlut <- rev(rainbow(1000, start = 0, end = 4/6)) # height color lookup table
joint_density_col_0 <- joint_density_colorlut[ floor(color_n * (plot_df_0$joint_density - joint_density_lim[1])/joint_density_range) + 1 ]

x1_lower_side <- 
  data.frame(
    x1 = rep(x1_lower, 101 + 2),
    x2 = c(x2_lower, seq(from = x2_lower, to = x2_upper, length = 101), x2_upper)
  )
x1_lower_side$z = c(0, dmvnorm(x1_lower_side[2:102, ], sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2)), 0)

x1_upper_side <- 
  data.frame(
    x1 = rep(x1_upper, 101 + 2),
    x2 = c(x2_lower, seq(from = x2_lower, to = x2_upper, length = 101), x2_upper)
  )
x1_upper_side$z = c(0, dmvnorm(x1_upper_side[2:102, ], sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2)), 0)

x2_lower_side <- 
  data.frame(
    x1 = c(x1_lower, seq(from = x1_lower, to = x1_upper, length = 101), x1_upper),
    x2 = rep(x2_lower, 101 + 2)
  )
x2_lower_side$z = c(0, dmvnorm(x2_lower_side[2:102, ], sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2)), 0)

x2_upper_side <- 
  data.frame(
    x1 = c(x1_lower, seq(from = x1_lower, to = x1_upper, length = 101), x1_upper),
    x2 = rep(x2_upper, 101 + 2)
  )
x2_upper_side$z = c(0, dmvnorm(x2_upper_side[2:102, ], sigma = matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2)), 0)

junk <- open3d()
z_max <- joint_density_lim[2]
plot3d(X1, X2, xlim=c(-3, 3), ylim=c(-3, 3), zlim=c(0, z_max), zlab="f(x1, x2)", xlab = "x1", ylab = "x2", mouseMode = "zAxis", type = "s")
surface3d(X1, X2, plot_df_0$joint_density, alpha = 0.3, col = joint_density_col_0)
surface3d(X1_solid, X2_solid, plot_df_0_solid$joint_density, alpha = 1, col = "grey")
polygon3d(x1_lower_side$x1, x1_lower_side$x2, x1_lower_side$z, coords = 2:3, col = "grey")
polygon3d(x1_upper_side$x1, x1_upper_side$x2, x1_upper_side$z, coords = 2:3, col = "grey")
polygon3d(x2_lower_side$x1, x2_lower_side$x2, x2_lower_side$z, coords = c(1, 3), col = "grey")
polygon3d(x2_upper_side$x1, x2_upper_side$x2, x2_upper_side$z, coords = c(1, 3), col = "grey")

#plotids <- with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
#                  type="s", col=as.numeric(Species)))
rglwidget(elementId = "plot_bivarnorm0_integral")
```

## Trivariate Normal

In three dimensions, we can't plot the density function directly - but we can visualize the distribution via its equi-density surfaces, which are analogous to the ellipses for the bivariate normal distribution.

Here's a plot of equi-density surfaces of the $\text{Normal}\left(\begin{bmatrix} 0 \\ 0 \\ 0 \end{bmatrix}, \begin{bmatrix} 1 & 0.3 & 0.6 \\ 0.3 & 2 & 1.1 \\ 0.6 & 1.1 & 3\end{bmatrix}\right)$ distribution.

```{r, echo = FALSE}
junk <- open3d()
plot3d(ellipse3d(x = matrix(c(1, .3, .6, .3, 2, 1.1, .6, 1.1, 3), nrow = 3), scale = rep(1, 3), centre = c(0, 0, 0), t = 3), type = "wire", col = "blue", alpha = 0.2, add = TRUE)
plot3d(ellipse3d(x = matrix(c(1, .3, .6, .3, 2, 1.1, .6, 1.1, 3), nrow = 3), scale = rep(1, 3), centre = c(0, 0, 0), t = 2), type = "wire", col = "yellow", alpha = 0.2, add = TRUE)
plot3d(ellipse3d(x = matrix(c(1, .3, .6, .3, 2, 1.1, .6, 1.1, 3), nrow = 3), scale = rep(1, 3), centre = c(0, 0, 0), t = 1), type = "wire", col = "red", alpha = 0.2, add = TRUE)
decorate3d()
rglwidget(elementId = "plot_trivarnorm")
```
